---
title: "Immunoprofiling of adenocarcinomas of the pancreatobiliary system"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About the data

These data represent a retrospective observational cohort study. The patient samples include patients who were diagnosed with adenocarcinoma arising in the pancreatobiliary system in Huddinge, Sweden. Tumor samples of hepatocellular carcinoma, a tumor type with a well-known immunohistochemical profile, were included as a control group for the results. 

I made some assumptions when analyzing these data. I assumed that the data on patient diagnosis were correct, and that they were collected and staged accordingly. I also assumed that the standards set by the authors (removing antibody markers with more than 40% missing values and removing tumor samples with more than 50% missing values) were logcially sound, as I did not change these parameters when tidying the data myself. Finally, I assumed that the tumor samples included in this study represent a normal distribution of data, with adequate sample size and equality of variance in standard deviation.


## Question 1: Are the staining markers well accounted for across the dataset? Can we create a new, tidy data frame for the immunohistochemical markers and tumor samples?

As in the paper, we will clean up the data by removing antibody markers (out of the initial 38) with more than 40% missing values as well as tumor samples with more than 50% missing values. We will also investigate if there are a significant number of missing or unknown cells, in which case we will impute these data. We will create this new tidy dataset as a new .csv file to include in our data folder and use for further analysis of future questions.

Import required packages
```{r}

library(plyr)
library(impute)
library(tidyverse)
BiocManager::install("impute")

```

Import raw dataset
```{r}
#read the raw data into a new variable

pcbilDataRaw <- read.csv(file = "C:/Users/jmalt/Desktop/tfcb_capstone/data/pcbil_raw.csv", row.names = 39, colClasses= c(rep("numeric",38), rep("character",2)), na.strings = "?",quote="'" )

# create working dataset for processing
pcbilDataWork <- pcbilDataRaw

# process column names

names(pcbilDataWork) <- tolower(names(pcbilDataWork))
names(pcbilDataWork)[39] <- "clin_diag"
pcbilDataWork$clin_diag <- as.factor(pcbilDataWork$clin_diag)

```

As they do in the paper, recode clinical diagnoses according to AJCC/IUCC-TNM 7 nomenclature, which is part of our tidying the data:
```{r}
# here I took cues from the paper to fix some diagnostic names
revalue(pcbilDataWork$clin_diag, c("Ampulla Ac" = "Ampullary carcinoma")) -> pcbilDataWork$clin_diag
revalue(pcbilDataWork$clin_diag, c("Distal Bile Duct Ac" = "Distal bile duct cancer")) -> pcbilDataWork$clin_diag
revalue(pcbilDataWork$clin_diag, c("Gallbladder Ac" = "Gallbladder cancer")) -> pcbilDataWork$clin_diag
revalue(pcbilDataWork$clin_diag, c("Hepatocellular Cancer" = "Hepatocellular carcinoma")) -> pcbilDataWork$clin_diag
revalue(pcbilDataWork$clin_diag, c("Intrahepatic Cholangiocarcinoma" = "Intrahepatic cholangiocarcinoma")) -> pcbilDataWork$clin_diag
revalue(pcbilDataWork$clin_diag, c("Pancreas Ac" = "Ductal pancreatic adenocarcinoma")) -> pcbilDataWork$clin_diag
revalue(pcbilDataWork$clin_diag, c("Perihilary Ac" = "Perihilar cholangiocarcinoma")) -> pcbilDataWork$clin_diag
```


Take a look at the composition of the raw dataset:
```{r}
#Total number of tumor samples
dim(pcbilDataWork)[1]

#Total number of markers
dim(pcbilDataWork)[2]

#List of markers
names(pcbilDataWork)
```


Composition of the raw dataset by atanomical-based diagnosis:
```{r}
# first, as they do in the paper, fix one incorrect anatomical-based diagnosis:
pcbilDataWork[rownames(pcbilDataWork) == "Intrahepatic Cholangiocarcinoma|366" , "clin_diag"] <- "Gallbladder cancer"

# then look at the table:
table(pcbilDataWork$clin_diag)
```

Summary of missing values by marker:
```{r}
# create a new variable to assess the extent of missing values
propmiss <- function(dataframe) {
  m <- sapply(dataframe, function(x) {
  	data.frame(
			nmiss=sum(is.na(x)), 
			n=length(x), 
			propmiss=round(sum(is.na(x))/length(x),2)
		)
	})
	d <- data.frame(t(m))
	d <- sapply(d, unlist)
	d <- as.data.frame(d)
	d$variable <- row.names(d)
	row.names(d) <- NULL
	d <- cbind(d[ncol(d)],d[-ncol(d)])
	return(d[order(d$propmiss), ])
}

reportmiss <- function(dataframe) {
  propMiss <- propmiss(dataframe)
  print(propMiss)
  totalNumValues <- dim(dataframe)[1] * (dim(dataframe)[2] - 1)
  totalMissingValues <- sum(propMiss$nmiss)
  return ( (100 / totalNumValues) * totalMissingValues  )
}

pcbilDataWorkMissingValues <- reportmiss(pcbilDataWork)

#Total percentage of missing data in processed dataset (%):
round(pcbilDataWorkMissingValues, 0)
```

Summary of missing values by tumor sample
```{r}
# as the above data show, look at the missing values according to tumor types

reportmissperrow <- function(dataframe) {

  pads <- c()
  numMissingValues <- c()
  propMissingValues <- c()
  
  aRow <- 1
  while(aRow <= nrow(dataframe)) {
    pads <- c(pads, row.names(dataframe)[aRow])
    numMissing <- sum(is.na(dataframe[aRow,]))
    propMissingValues <- c(propMissingValues, round((100 / (dim(dataframe)[2] -1)) *numMissing,2))
  	numMissingValues <- c(numMissingValues, numMissing)
  	aRow <- aRow + 1
  }
  missingD <- data.frame(pad = pads, nmissing = numMissingValues, propmissing = propMissingValues, stringsAsFactors=F)
  missingD$propmiss_interv <- cut(missingD$propmissing, c(0,10,20,30,40,50,60,70,80,90,100))
  
  print(table(missingD$propmiss_interv))
  
  return (subset(missingD, propmissing > 50)[, 1])
}

numCasesMissGreat50 <- reportmissperrow(pcbilDataWork)
#Number of cases in work dataset with more than 50% missing data:
length(numCasesMissGreat50)
```

The original paper removes markers with > 40% missing values (as determined by the output above, under "Summary of missing values by marker") and these are:
ttf1, cd146, cd146_nucl, ngfr_str, calretinin, glypican_3, ezh2, synap, hbme1, mesothelin, hepatocyte
```{r}
pcbilDataMarkerTrimmed <- subset(pcbilDataWork, select=-c(ttf1, cd146, cd146_nucl, ngfr_str, calretinin, glypican_3, ezh2, synap, hbme1, mesothelin, hepatocyte))

pcbilDataMarkerTrimmedMissingValues <- reportmiss(pcbilDataMarkerTrimmed)

```


Missing values in marker trimmed dataset, by tumor sample:
```{r}
samplesWithMissingGreater50 <- reportmissperrow(pcbilDataMarkerTrimmed)
#Number of cases in filtered dataset with more than 50% missing data:
length(samplesWithMissingGreater50)
print(samplesWithMissingGreater50)

```

Next, as in the original paper, we remove tumor samples with > 50% missing values:
```{r}
pcbilDataMarkerSamplesTrimmed <- pcbilDataMarkerTrimmed[!rownames(pcbilDataMarkerTrimmed) %in% samplesWithMissingGreater50, ]

pcbilDataMarkerTrimmedMissingValues <- reportmiss(pcbilDataMarkerSamplesTrimmed)

reportmissperrow(pcbilDataMarkerSamplesTrimmed)

```


Next the paper imputes the missing values using KNN-based imputation:
```{r}
pcbilMatrixImputed <- impute.knn(as.matrix(pcbilDataMarkerSamplesTrimmed[1:27]))
pcbilDataImputed <- as.data.frame(pcbilMatrixImputed$data)
pcbilDataImputed$clin_diag <- pcbilDataMarkerSamplesTrimmed$clin_diag

# check the result:

pcbilDataImputedMissingValues <-reportmiss(pcbilDataImputed)

```

Finally we save the imputed, tidy dataset:
```{r}
pcbilDataImputed$pad <- rownames(pcbilDataImputed)
write.csv(pcbilDataImputed, 'data/pcbilDataImputed.csv', row.names=F, na="")
```


Interpretation of results, answering question

## Question 2: Of the 5 markers with the highest average staining intensity, which tumor type has the highest average staining intensity for each one?

```{r}
library(dplyr)

#read the new imputed tidy dataset into a new variable

pcbilDataImputed <- read.csv(file = "C:/Users/jmalt/Desktop/tfcb_capstone/data/pcbilDataImputed.csv")

df <- subset(pcbilDataImputed, select = -c(pad,clin_diag)) %>%
  print()


```
```{r}
names(df)
```


```{r}

# calculate the mean intensity of each marker

col_avg <- colMeans(df) %>%
  as.data.frame() %>%
  print()
```

```{r}
# create a new dataframe with the average intesities of each marker

df_col_avg <- data.frame(marker=c('ck5','ck7','ck17','ck18','ck19','ck20','vim','muc1','muc2','muc5ac','muc6','berep4','ema','mcea','pcea','ca125','ca19.9','maspin','wt1cyt','cdx2','p53','p63','ki67','smad4','chra','cd56','cd10'), avg_intensity=c(4.0118465,65.2483700,19.6879293,89.9241933,76.1899959,10.0508693,19.8194954,49.1249830,3.2207426,29.9493465,9.7273863,63.6890499,72.1461365,29.9998545,73.1693469,29.0709046,53.9108162,48.0657915,7.6755225,14.4455777,30.9228179,2.8586118,33.8746895,60.1012852,0.9427465,4.5393405,12.0211105))
```


```{r}
# find the 5 markers with highest average staining intensity

df_col_avg_sorted_asc <- df_col_avg[order(-df_col_avg$avg_intensity),] %>%
  head(,5) %>%
  print()

```
```{r}
# calculate the mean of each of the 5 previous markers for each separeate tumor type

df_refined <- pcbilDataImputed%>%
  select(ck18, ck19, pcea, ema, ck7, clin_diag)%>%
  group_by(clin_diag) %>%
  summarise_at(1:5, ~ round(mean(.), 5))%>%
  print()
```

```{r}
results <- as.data.frame(df_refined)%>%
  print()
```

```{r}
# help with this section from: https://www.r-bloggers.com/2016/11/make-a-bar-plot-with-ggplot/

# make a new variable that has the info above, ready to graph with ggplot2

marker <-rep(c("ck18", "ck19", "pcea", "ema", "ck7"), 7)
Ampullary_carcinoma <-c(90.25236, 91.97917, 78.31481, 84.40104, 59.09375)
Distal_bile_duct_cancer <-c(81.42857, 92.50000, 89.08730, 92.50000, 81.78571)
Ductal_pancreatic_adenocarcinoma <-c(90.73342, 91.73291, 85.47480, 89.98884, 82.25719)
Gallbladder_cancer <-c(91.94726, 90.09760, 74.86149, 83.88983, 75.57432)
Hepatocellular_carcinoma <-c(87.87500, 12.13746, 45.37776, 17.00627,  9.23718)
Intrahepatic_cholangiocarcinoma <-c(89.70511, 90.52062, 72.67535, 80.36352, 81.06529)
Perihilar_cholangiocarcinoma <-c(91.60365, 92.40741, 80.86111, 77.79707, 69.70370)
values <-c(Ampullary_carcinoma, Distal_bile_duct_cancer, Ductal_pancreatic_adenocarcinoma, Gallbladder_cancer, Hepatocellular_carcinoma, Intrahepatic_cholangiocarcinoma, Perihilar_cholangiocarcinoma)
Tumor <-c(rep("Ampullary carcinoma", 5), rep("Distal bile duct cancer", 5), rep("Ductal pancreatic adenocarcinoma",5), rep("Gallbladder cancer", 5), rep("Hepatocellular carcinoma", 5), rep("Intrahepatic cholangiocarcinoma", 5), rep("Perihilar cholangiocarcinoma", 5))
results_graph <- data.frame(marker, values)

```

```{r}
# use ggplot to graph results

library(ggplot2)

p <-ggplot(results_graph, aes(marker, values))
p +geom_bar(stat = "identity", aes(fill = Tumor), position = "dodge", width = 0.8) +
  xlab("Marker") + ylab("Staining intensity") +
  ggtitle("Immunohistochemistry of strongest markers in PCBIL") +
  theme_bw() +
  theme(aspect.ratio=4/7)

```


## Reproducibility

The analysis in the original paper, at least in terms of "tidying" the data, was relatively easy to reproduce. It was easy to understand the workflow in R, especially when it came to pruning the data set according to their standards (ie. removing antibody markers with more than 40% missing values and tumor samples with more than 50% missing values). The original paper however only provided the raw data file, and not the tidy data, so I have to assume that I worked through their analysis correctly. Ultimately, I am satisfied with my tidy data file ('data/pcbilDataImputed.csv'). I have included code comments to help a prospective user work through my own analysis, including through Question 2. I will document the structure of this repo in the README file in github (as well as README files in each subsequent directory), so that others can reliably reproduce this analysis. 